# Pre-processing of DNA methylation data

In this chapter will explain the steps for the pre-processing of
DNA methylation data with the meffil R package.

## Creation of the Sample Sheet
After loading the meffil library, we need to create a sample sheet from the IDAT files.

```{r ss, message=FALSE, warning=FALSE, results=FALSE}
library(meffil)
samplesheet <- meffil.create.samplesheet("H:/analyses/Course_Intro_to_EWAS/db/GSE42861/Raw/G1", recursive=TRUE)
```
```{r ssview}
dim(samplesheet)
```

The function automatically adds the sample name, slide, row, column and filename
from the IDATs. We can not remove these columns but we can manually add more.
In this case we add the sex, age and smoking status variables. If we had duplicated
samples for the same individual, we would also add the individual ID and then 
remove the duplicates. 

First we load the phenotype data
```{r phenoload}
pheno <- read.table("H:/analyses/Course_Intro_to_EWAS/db/GSE42861/Phenotype/GSE42861_pheno.txt", sep = '\t', header = T)
head(pheno)
```

And then add the sex, the age and the smoking status variables, keeping only
those samples with metadata
```{r phenoadd}
samplesheet <- merge(samplesheet[,c('Sample_Name', 'Slide', 'sentrix_row', 'sentrix_col', 'Basename')], 
                     pheno[,c('Sample', 'Age', 'Sex', 'Smoking')], 
                     by.x = 'Sample_Name', 
                     by.y = 'Sample')
head(samplesheet)
```

Finally we can save the sample sheet
```{r save, eval=FALSE}
save(samplesheet, file="H:/analyses/Course_Intro_to_EWAS/results/QC/SampleSheetG1.Rdata")
```

## Sample and CpG quality control
Once we have the sample sheet, the first step for the QC is the creation of the 
QC objects. We specify the cell type reference (idol optimized)
and the thresholds for detection and minimum number of beads. 

**DO NOT RUN THIS STEP IN THE CLASS** - it is very time consuming. 
```{r objects, eval=FALSE}
# DO NOT RUN

meffil.list.cell.type.references()
qc.objects <- meffil.qc(samplesheet, cell.type.reference="blood idoloptimized", 
                        verbose=TRUE,
                        detection.threshold	= 0.01,
                        bead.threshold = 3)
save(qc.objects, file = 'H:/analyses/Course_Intro_to_EWAS/results/QC/QCObjects.RData')
```
Instead, load the QC Objects that we provide:
```{r loadqc}
load('H:/analyses/Course_Intro_to_EWAS/results/QC/QCObjects.RData')
```

After that, we setup the parameters for the QC
```{r param}
qc.parameters <- meffil.qc.parameters(
  beadnum.samples.threshold             = 0.05,
  detectionp.samples.threshold          = 0.03,
  detectionp.cpgs.threshold             = 0.05, 
  beadnum.cpgs.threshold                = 0.05,
  sex.outlier.sd                        = 3
)
```

And run it
```{r summary, warning=FALSE, message=FALSE, results=FALSE}
qc.summary <- meffil.qc.summary(
  qc.objects,
  parameters = qc.parameters,
  genotypes=NULL,
  verbose=TRUE
)
```

To generate the QC report we can run:
```{r qcrep, eval=FALSE}
meffil.qc.report(qc.summary, output.file="H:/analyses/Course_Intro_to_EWAS/results/QC/qc-report_whole.html")
```
This line creates an html report with several QC tests. We need to examine it
with detail.


The next step is getting rid of the samples that do not pass the QC. In this
case, after looking at the report, we have decided to maintain those samples
that are outliers for the XY ratio.
```{r outliers}
outlier <- qc.summary$bad.samples
outlier <- outlier[outlier$issue!='X-Y ratio outlier',]
outlier
qc.objects <- meffil.remove.samples(qc.objects, outlier$sample.name)
```
The probes that do not pass the QC will be removed afterwards during the
normalisation step.

## Normalisation
First of all we assess the number of needed PCs with the help of this plot:
```{r fitpcs, warning=FALSE, message=FALSE, results=FALSE}
y <- meffil.plot.pc.fit(qc.objects)
y$plot
```

Then we perform the functional normalisation indicating the number of PCs and 
the IDs of bad CpGs from the QC to remove. After this step it is recommended to check
again the missingness per sample and cpg

**DO NOT RUN THOSE STEPS IN THE CLASS** - they are very time consuming. 
```{r norm, eval=FALSE}
# DO NOT RUN
norm.objects <- meffil.normalize.quantiles(qc.objects, number.pcs=15)
# DO NOT RUN
norm.beta <- meffil.normalize.samples(
  norm.objects,
  just.beta=T,
  remove.poor.signal = T,
  cpglist.remove=qc.summary$bad.cpgs$name,
  verbose=T)

# missing per sample - DO NOT RUN
miss_IDs <- miss_case_summary(data.frame(t(norm.beta)), order = TRUE)
miss_IDs$case <- colnames(norm.beta)
gc(reset=TRUE)
bad_ids <- miss_IDs[which(miss_IDs$pct_miss>5),]$case

# missing per cpg - DO NOT RUN
bad_cpgs <- miss_cpgs[which(miss_cpgs$pct_miss>5),]$variable
length(bad_cpgs)

# remove - DO NOT RUN
norm.beta <- norm.beta[!rownames(norm.beta) %in% bad_cpgs, !colnames(norm.beta) %in% bad_ids]
norm.objects <- norm.objects[!names(norm.objects)%in% bad_ids]
gc(reset=TRUE)

save(norm.objects, file = "H:/analyses/Course_Intro_to_EWAS/results/QC/norm_objects15.RData")
save(norm.beta, file = "H:/analyses/Course_Intro_to_EWAS/results/QC/norm_beta15.Robj")
```
Instead, load the normalisation objects and the betas that we provide:
```{r load_norm}
load("H:/analyses/Course_Intro_to_EWAS/results/QC/norm_objects15.RData")
load("H:/analyses/Course_Intro_to_EWAS/results/QC/norm_beta15.Robj")
```

Now we can create a normalisation report. We need to convert to factor all the
batch variables that we want to assess and recalculate the PCs
```{r pre_norm, message=FALSE, results=FALSE}
for (i in 1:length(norm.objects)){
  norm.objects[[i]]$samplesheet$Slide<-as.factor(norm.objects[[i]]$samplesheet$Slide)
  norm.objects[[i]]$samplesheet$Sex<-as.factor(norm.objects[[i]]$samplesheet$Sex)
  norm.objects[[i]]$samplesheet$Age<-as.factor(norm.objects[[i]]$samplesheet$Age)
  norm.objects[[i]]$samplesheet$Smoking<-as.factor(norm.objects[[i]]$samplesheet$Smoking)
}

batch_var<-c("Slide","Sex", "Age", "Smoking")

norm.parameters <- meffil.normalization.parameters(
  norm.objects,
  variables=batch_var,
  control.pcs=1:5,
  batch.pcs=1:5,
  batch.threshold=0.01
)

pcs <- meffil.methylation.pcs(norm.beta,verbose=T, full.obj = T)
```

Finally, we obtain the normalisation summary and report
```{r normsumm, eval=FALSE}
norm.summary <- meffil.normalization.summary(norm.objects, pcs$x, parameters=norm.parameters)
meffil.normalization.report(norm.summary, output.file="H:/analyses/Course_Intro_to_EWAS/results/QC/normalization-report.html")
```
## Creation of a Expression Set

### Estimation of cell type proportions

To run the EWAS, we will need to know the cell type proportions. They are
automatically calculated in the first steps of the QC and can be recovered and
saved this way:
```{r cc}
cc<-t(sapply(qc.objects, function(obj) obj$cell.counts$counts))
cc<-data.frame(PatientID=row.names(cc),cc) 
write.table(cc, file = "H:/analyses/Course_Intro_to_EWAS/results/QC/cellular_composition.txt", quote = F, row.names = F, sep = '\t')
```

We add this information to the phenotype that we already have
```{r phenodata}
phenoData <- merge(pheno, cc[,2:7], by.x = 'Sample', by.y = 0)
rownames(phenoData) <- phenoData[,1]
phenoData <- phenoData[,-1]
head(phenoData)
```


### Assembling

To assemble the final ExpressionSet we first create the metadata for the
phenotypes, which consists on a dataframe with the variables description.
Then we create an Annotated Dataframe. 

```{r mdata, message=FALSE, results=FALSE}
library(Biobase)
mData <- data.frame(labelDescription=c("Sample cell type","Patient Age", "Patient Sex", "Smoking Status (never/current/former)",
                                       "Percentage of B Lymphocytes", "Percentage of CD4T Lymphocytes",
                                       "Percentage of CDBT Lymphocytes", "Percentage of Monocytes", "Percentage of Neutrophils",
                                       "Percentage of Natural Killer"),
                    row.names=colnames(phenoData))

phenoData = new("AnnotatedDataFrame", data=phenoData, varMetadata=mData)
```

After that, we join the matrix of the normalized betas with the annotated
dataframe of the phenotypes and cellular compositions.
We have to make sure that the sample IDs coincide in the two datasets.
```{r eset}
all(colnames(norm.beta) == rownames(phenoData))
eSet<-ExpressionSet(assayData=norm.beta,
                    phenoData=phenoData)
eSet
```

Once we have saved the ExpressionSet with the normalised betas and the
phenotypes we are ready to run the EWAS.

**DO NOT RUN THIS STEP IN THE CLASS** - it is very time consuming. We will
provide the eSet needed for the EWAS.
```{r saveeset, eval=FALSE}
# DO NOT RUN
saveRDS(eSet, 'H:/analyses/Course_Intro_to_EWAS/results/QC/GSE42861_full_norm.Rds')
```

