# Meta-analysis of epigenome-wide association studies (metaEWAS)

In this chapter we will run a meta-analysis of epigenome-wide association studies (metaEWAS).
A file with the theory can be downloaded [here](./Theory/ISG_summer_school_EWAS4.pptx)


## Install and load packages
Today we will load these packages

## Quality control of results from different studies
## Statistical methods to combine results from different studies
## Visualization of results: QQ-plot, Manhattan, Volcano
# Epigenome-wide association study (EWAS)
In this chapter we will run an Epigenome-Wide Association Study (EWAS)

## Install and load packages
Today we will load these packages

```{r librariesload}
library(data.table)
library(qqman)
library(metafor)
library(reshape)
library(plyr)
#library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(ggplot2)# plots
library(ggrepel)
library(GenomicRanges) # prepare db for Manhattan plot
library(karyoploteR)
```


## Load EWAS results by cohort

Here we will show an example to run the metaEWAS testing never smokers against 
current smokers. Nevertheless, the same code can be applied to run other
analyses testing never vs former or current vs former. 

We begin loading the Never vs Current EWAS results from 3 cohorts: 
cohort1, cohort2 and cohort3. 
```{r load_metaewas, eval=FALSE}
EWASres.cohort1 <- read.table(file = "Results.EWAS.cohort1.NeverVScurrent.txt",sep=",",header = TRUE)
EWASres.cohort2 <- read.table(file = "Results.EWAS.cohort2.NeverVScurrent.txt",sep=",",header = TRUE)
EWASres.cohort3 <- read.table(file = "Results.EWAS.cohort3.NeverVScurrent.txt",sep=",",header = TRUE)
```

## Quality control of EWAS results from the different studies

Before running the metaEWAS, it is really important to check the quality of the
EWAS results that we want combine through a meta-analysis. 
This includes: 
- examining inflation
- create a beta's box plot
- create a precision plot 

```{r MAfolder, eval=FALSE}
#Set wd

#res.dir<-"V:/analyses/Course_Intro_to_EWAS/results"
#setwd(res.dir)

#Create the directory to save EWAS results
dir.create("MA.results")

setwd(paste0(res.dir,"/MA.results"))
dir.create("QC")
setwd(paste0(res.dir,"/MA.results/QC"))
```
### Lambda and qqplot
We look to the lambda and the QQplot to examine inflation
```{r lambdas, eval=FALSE}
lambda.cohort1<- qchisq(median(EWASres.cohort1$p.value,na.rm=T), df = 1, lower.tail = F)/qchisq(0.5, 1) 
lambda.cohort2<- qchisq(median(EWASres.cohort2$p.value,na.rm=T), df = 1, lower.tail = F)/qchisq(0.5, 1)
lambda.cohort3<- qchisq(median(EWASres.cohort3$p.value,na.rm=T), df = 1, lower.tail = F)/qchisq(0.5, 1)

lambdas<-cbind(lambda.cohort1,lambda.cohort2,lambda.cohort3)
lambdas
write.table(lambdas,"Lambdas.NeverVScurrent.txt",sep=",")
```
```{r QQlots, eval=FALSE}
#cohort 1
pvals.cohort1<-EWASres.cohort1$p.value
qq(pvals.cohort1,main=("QQPlot EWAS cohort 1 smoking Never VS Current")) 
 
#cohort 2
pvals.cohort2<-EWASres.cohort2$p.value
qq(pvals.cohort2,main=("QQPlot EWAS cohort 2 smoking Never VS Current")) 
 
#cohort 3
pvals.cohort3<-EWASres.cohort3$p.value
qq(pvals.cohort3,main=("QQPlot EWAS cohort 3 smoking Never VS Current")) 
```
### Box plot and Precision plot

We will create a beta’s box plot and a precision plot comparing the different EWAS results

First, we will prepare a dataframe with the beta information of the three cohorts.
```{r BoxPlotdf, eval=FALSE}
#create df for Box plot

QCPlot_cohort1<-EWASres.cohort1[,c("probeID","coefficient")]
QCPlot_cohort1$Cohort<-"cohort1"
head(QCPlot_cohort1)

QCPlot_cohort2<-EWASres.cohort2[,c("probeID","coefficient")]
QCPlot_cohort2$Cohort<-"cohort2"
head(QCPlot_cohort2)

QCPlot_cohort3<-EWASres.cohort3[,c("probeID","coefficient")]
QCPlot_cohort3$Cohort<-"cohort3"
head(QCPlot_cohort3)

QCPlot.df<-rbind(QCPlot_cohort1,QCPlot_cohort2,QCPlot_cohort3)
```
Beta's Box plot
```{r BoxPlot, eval=FALSE}
p1<-ggplot(data=QCPlot.df,aes(x=Cohort, y=coefficient, fill=Cohort)) +
  geom_boxplot() 
p1
```
Precision plot

By plotting 1 divided by the median of the effect SE against the square root of the sample size for each cohort.
```{r PrecisionPlotdf, eval=FALSE}
invSE_cohort1<-1/median(EWASres.cohort1$coefficient.se)
invSE_cohort2<-1/median(EWASres.cohort2$coefficient.se)
invSE_cohort3<-1/median(EWASres.cohort3$coefficient.se)
                        

precision <- data.frame(
  Cohort = c("cohort1","cohort2","cohort3"),
  N = c(188, 176, 166),
  invSE= c(invSE_cohort1,invSE_cohort2,invSE_cohort3))
  
precision$sqrt_N<-sqrt(precision$N)
head(precision)
```

```{r PrecisionPlot, eval=FALSE}
p2<-ggplot(data=precision,aes(x=sqrt_N, y=invSE, color=Cohort)) +  geom_point() 
p2
```

## Statistical methods to combine results from different studies

We will combine the results through a fixed-effect weight meta-analysis using Metafor R package

### Data preparation

We want to create a dataframe merging the EWAS results from the three cohorts
(probeID, coefficient, se, p.value, n)

First, we subset the EWAS results to probeID, coefficient, se .p.value and n
and we change the name of the column in order to merge these results in a dataframe

```{r subsetEWASresults, eval=FALSE}
# sample_list <- c("Results.EWAS.cohort1.NeverVScurrent.txt","Results.EWAS.cohort2.NeverVScurrent.txt","Results.EWAS.cohort3.NeverVScurrent.txt")
# 
# translation.table <- matrix(c("probeID", "probeID","probeID", "BETA","BETA","BETA", "SE", "SE","SE", "P_VAL", "P_VAL","P_VAL"), ncol=2, byrow=TRUE)
# write.table(translation.table, "translation.table.txt")

# EWASres.cohort1$N<-188
# EWASres.cohort1$N<-176
# EWASres.cohort1$N<-166

EWASres.cohort1<-EWASres.cohort1[,c("probeID","coefficient", "coefficient.se","p.value","n")]
EWASres.cohort2<-EWASres.cohort2[,c("probeID","coefficient", "coefficient.se","p.value","n")]
EWASres.cohort3<-EWASres.cohort3[,c("probeID","coefficient", "coefficient.se","p.value","n")]
#sample_N <- data.frame(file = sample_list, N = c(188,176,166), stringsAsFactors = FALSE)
names(EWASres.cohort1)<- c("probe", "coef_cohort1", "se_cohort1", "P_VAL_cohort1", "N_for_probe_cohort1")
names(EWASres.cohort2)<- c("probe", "coef_cohort2", "se_cohort2", "P_VAL_cohort2", "N_for_probe_cohort2")
names(EWASres.cohort3)<- c("probe", "coef_cohort3", "se_cohort3", "P_VAL_cohort3", "N_for_probe_cohort3")
```

Second, we merge the EWAS results of the 3 cohorts
```{r mergeEWASresults, eval=FALSE}
dat <- merge(EWASres.cohort1, EWASres.cohort2, by="probe", all.x=TRUE, all.y=TRUE)
dat <- merge(dat, EWASres.cohort3, by="probe", all.x=TRUE, all.y=TRUE)
dim(dat)
head(dat)
```
Then, we create a new column with the total N of the metaEWAS and remove those 
probes present in <50% of the cohorts

```{r totalN, eval=FALSE}
dat$N <- rowSums(dat[,c("N_for_probe_cohort1", "N_for_probe_cohort2", "N_for_probe_cohort3")], na.rm=TRUE)
totN = sum(max(dat$N_for_probe_cohort1, na.rm=TRUE), max(dat$N_for_probe_cohort2, na.rm=TRUE), max(dat$N_for_probe_cohort3, na.rm=TRUE))
datN <- dat[,c("probe", "N_for_probe_cohort1", "N_for_probe_cohort2", "N_for_probe_cohort3")]
datN <- datN[which(rowMeans(!is.na(datN[-1]))>=0.5),]
dat <- dat[which(dat$probe %in% datN$probe),]
dat<-dat[which(dat$N >= totN/2),]
dat <- droplevels(dat)
studies <- c("cohort1","cohort2","cohort3")
head(dat)

```

### Run fixed effect meta-analysis

```{r fixed.effects.meta.analysisFunction, eval=FALSE}
fixed.effects.meta.analysis <- function(list.of.studies,data){
  require(metafor)
  coefs = data[,c("probe",paste0("coef_",list.of.studies))]
  ses = data[,c("probe",paste0("se_",list.of.studies))]
  require(reshape)
  coefs = melt(coefs)
  coefs = cbind(coefs, colsplit(coefs$variable, "_", names = c("var", "study")))
  coefs <- coefs[,c("probe","study","value")]
  names(coefs) <- c("probe","study","coef")
  ses = melt(ses)
  data.long = cbind(coefs,ses[,"value"])
  names(data.long)<-c("probe","study","coef","se")
  res = split(data.long, f=data$probe)
  res = lapply(res, function(x) rma.uni(slab=x$study, yi=x$coef,sei=x$se,method="FE",weighted=TRUE))
  res
}
```

```{r extract.and.merge.meta.analysisFuntion, eval=FALSE}
extract.and.merge.meta.analysis <-function(meta.res,data){
  require(plyr)
  data.meta = ldply(lapply(meta.res, function(x) unlist(c(x$b[[1]],x[c("se","pval","QE","QEp","I2","H2")]))))
  colnames(data.meta)<-c("probe","coef.fe","se.fe","p.fe","q.fe","het.p.fe","i2.fe","h2.fe")
  data = merge(data,data.meta,by="probe",all.x=T)
  data
}
```

To run the Never vs Current metaEWAS, we will use tha above functions (without edition)

- fixed.effects.meta.analysis function
- extract.and.merge.meta.analysis function

```{r RunMetaEWAS, eval=FALSE}
meta.results.current <- fixed.effects.meta.analysis(list.of.studies = studies, data = dat)
dat.current <- extract.and.merge.meta.analysis(meta.res = meta.results.current, data = dat)
dat.current$fdr<-p.adjust(dat.current$p.fe, method = "fdr")
head(dat.current)
```

### Explore the results

We order the results by the metaEWAS p.value
```{r order, eval=FALSE}
dat.current.ord<-dat.current[order(dat.current$p.fe),]
dim(dat.current.ord)

write.table(xxxx)
```

Significant hits 
Our results were corrected for multiple testing using FDR method. Significance 
was defined at FDR p-value < 0.05. 

```{r significant, eval=FALSE}
sig.current<-dat.current.ord[dat.current.ord$fdr <0.05,]
dim(sig.current)
head(sig.current)
```
Lambda and QQplot
```{r lambda, eval=FALSE}
lambda.current<- qchisq(median(dat.current.ord$p.fe,na.rm=T), df = 1, lower.tail = F)/qchisq(0.5, 1)
lambda.current 
```

```{r qqplot, eval=FALSE}
pvals.current<-dat.current.ord$p.fe
qq(pvals.current,main=("QQPlot metaEWAS smoking Never VS Current")) 
```

Volcano plot
```{r VolcanoPlot, eval=FALSE}
dat.current.ord$diffexpressed <- "NO"
dat.current.ord$diffexpressed[dat.current.ord$coef.fe  > 0 & dat.current.ord$fdr <0.05] <- "POSITIVE"
dat.current.ord$diffexpressed[dat.current.ord$coef.fe  < 0 & dat.current.ord$fdr <0.05] <- "NEGATIVE"

p <- ggplot(data=dat.current.ord, aes(x=dat.current.ord$coef.fe, y=-log10(dat.current.ord$p.fe), col=dat.current.ord$diffexpressed)) +
  xlim(c(-0.3,0.3))+ ylim(c(0,130)) + 
  geom_point(size = 2) + theme_minimal() +
  labs(title = " ", x = "beta", y = "-log10(P-value)", colour = "Effect") + 
  theme(axis.title = element_text(family = "Helvetica", size = 14, color = "black",vjust=0.5)) + 
  theme(plot.title = element_text(family = "Helvetica",size=14,face="bold",color="black", hjust= 0.5, vjust=0.5)) + 
  theme(legend.title = element_text(family = "Helvetica", color = "black", size = 14))

p2 <- p + geom_hline(yintercept=c(-log10(0.012)), col=c("red"),linetype = "dashed") + 
  theme(axis.text = element_text(size = 14))
mycolors<-c("#157F8D","#AF8D9B", "grey")
names(mycolors) <- c("POSITIVE", "NEGATIVE", "NO")

p3 <- p2 + scale_colour_manual(values = mycolors)

p3
```
Forest plot
```{r ForestPlot, eval=FALSE}
#Forest plots

X<-"cg05575921"
G<-"AHRR"
I2<-dat.current.ord$i2.fe
forest(meta.results.current$cg05575921,digits=4,
       xlab=expression(paste("Change in DNA methylation never vs current")),
       mlab="Fixed-effects meta-analysis",col="deepskyblue2",border="deepskyblue2",cex=0.8)
#title(paste0("Forest plot for probe:\n", X," (",G,"), I-squared: ",I2,""),line=-2)
```
