# Biological interpretation of findings
In this chapter we will explain the steps for the biological interpretation of
the findings. We will need to load some libraries:
```{r libs_annot, eval=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
```


## Annotation of CpG sites
Once we have obtained a list with the CpGs that are significant in our analysis,
we need to locate them in the genome and try to know which structures 
surroud them. The annotation consists in obtaining this information:
```{r annot}

```

## Functional enrichment analysis
The idea is to compare the list of genes that overlap our CpGs with the list of 
all the human genes that are anotated in certain databases. With this, we can see
if our list of genes are enriched in any kind of function.

First we need to obtain the list of all human genes that are curated in specific
databases. We will work with the Gene Ontology (GO) and the Kyoto Encyclopedia of
Genes and Genomes (KEGG) databases
```{r genelist, eval=FALSE}
# GO
df = as.data.frame(org.Hs.egGO)
go_gene_list = unique(sort(df$gene_id))

# KEGG
dfk = as.data.frame(org.Hs.egPATH)
kegg_gene_list = unique(sort(dfk$gene_id))
```


And run the analyses for the
```{r go, eval=FALSE}
ans.go <- enrichGO(gene = entrez, ont = "BP",
                   OrgDb ="org.Hs.eg.db",
                   universe = go_gene_list,
                   readable=TRUE,
                   pvalueCutoff = 0.05)

tab.go <- as.data.frame(ans.go)
tab.go<- subset(tab.go, Count>5)

```

## Comparison with EWAS Atlas and EWAS Catalog